Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-08 appendix.tex
updating appendix

\section{Limitations of the vertically global shearing box}\label{global_corr}
Here, we discuss the relation between the linear problem in the
VGSB framework and the global linear problem.

 This procedure produces $\phi$ and $z$ momentum equations
that are equilvalent to those for $y$ and $z$ momenta in the VGSB.

Using the notation and disk models in the main text, the
energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_rc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_r  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_r + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_r & = \ii k_r W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_\phi,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$. To obtain the
linear problem in the VGSB, we take the radially-local 
approximation further and formally set $\hat{g}_c=0$, $r\to x$ and
$k_r\to k_x$. It is clear that this approximation improves with
increasing $|k_x|$.

@@

\subsection{Spurious growth of adiabatic perturbations in the local
  model with vertical shear}\label{analytic_adia} 
For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
simply use the Soilberg-Hoiland critera to assess stability
(\S\ref{solberg}). Nevertheless, we briefly consider this case in the
local framework to demonstrate its limitations. For $\beta\to\infty$
the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
be written as

\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat \notag\\
  &=  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
  &\phantom{=} +
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat\notag\\
  &\phantom{=} + \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear.

>>>>

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we used the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytical progress for vertically
isothermal disks. Here, we discuss limitations of this approach.

The main difference between  VGSB framework

Using the notation and disk models in the main text, the
energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_rc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_r  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_r + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_r & = \ii k_r W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_\phi,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$.

%  To obtain the
% linear problem in the VGSB, we take the radially-local 
% approximation further and formally set $\hat{g}_c=0$, $r\to x$ and
% $k_r\to k_x$. It is clear that this approximation improves with
% increasing $|k_x|$.

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
simply use the Soilberg-Hoiland critera to assess stability
(\S\ref{solberg}). Nevertheless, we briefly consider this case in the
local framework to demonstrate its limitations. For $\beta\to\infty$
the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
be written as

\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear.

@@

### 2015-02-08 numerical.tex
updating appendix

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI mode as a function of $\beta$ for $\khat=30$. This is
approximately the perturbation wavenumber observed in the fiducial
numerical simulation of \cite{nelson13}. We also plot the growth rates
obtained by numerically solving the dispersion relation 
Eq. \ref{relax_disp} and the upper limit to 
the thermal relaxation timescale $\beta_\mathrm{crit}$ given by
Eq. \ref{iso_vsi_cond}. For the present disk parameters
$\beta_\mathrm{crit}=0.125$. We find good agreement between the
numerical and analytical growth rates, as well as the critical
relaxation timescale. Growth rates are insensitive to $\beta$ for
$\beta\lesssim 0.05$, beyond which growth rates rapidly decrease.

For this disk model, \citeauthor{nelson13} found a thermal relaxation
timescale of $\beta\gtrsim 0.6$ stabilized the VSI in non-linear
hydrodynamic simulations. This is consistent with our numerical and
analytic results.

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to the
critical value beyond which the fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Increasing $\beta$
restricts  the region in which $W\sim z$ closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important.

@@

\subsection{Critical thermal relaxation timescale}
We test our estimate for the maximum thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental
VSI (Eq. \ref{iso_vsi_cond}). We use the previous disk
parameters as a fiducial case and vary $q\in[-0.2,-1.2]$,
$\gamma\in[1.2,2.0]$ and $\epsilon\in[0.02,0.1]$
separately. The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.

We find Fig. \ref{bcrit_compare} generally agrees with
Eq. \ref{iso_vsi_cond}.  The match is better for smaller
$\beta_\mathrm{crit}$. There is a noticeable difference as
$\gamma\to1$ because the  derivation of $\beta_\mathrm{crit}$ assumed
$\gamma>1$.

>>>>

Fig. \ref{bcrit_compare1} shows the growth rate $\nu$ of the
fundamental VSI mode as a function of $\beta$ for $\khat=30$. This is
approximately the perturbation wavenumber observed in the fiducial
numerical simulation of \cite{nelson13}. We also plot the growth rates
obtained by numerically solving the dispersion relation 
Eq. \ref{relax_disp} and the upper limit to 
the thermal relaxation timescale $\beta_\mathrm{crit}$ given by
Eq. \ref{iso_vsi_cond}. For the present disk parameters
$\beta_\mathrm{crit}=0.125$.

We find good agreement between the numerical and analytical growth
rates, as well as the critical relaxation timescale. Growth rates are
insensitive to $\beta$ for $\beta\lesssim 0.05$, beyond which growth
rates rapidly decrease. For this disk model, \citeauthor{nelson13}
found a thermal relaxation timescale of $\beta\gtrsim 0.6$ stabilized
the VSI in non-linear hydrodynamic simulations. This is consistent
with our results.

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to the
critical value beyond which the fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Increasing $\beta$ 
restricts  the region in which $W\sim z$ closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important. Finite thermal relaxation increases $|W|$ near the
vertical boundaries relative to that about the midplane, but
perturbations near the disk surface is unlikely important in practice
because the disk atmosphere contains little mass.

@@

\subsection{Critical thermal relaxation timescale}
% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We test our estimate for the maximum thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental
VSI (Eq. \ref{iso_vsi_cond}). We use the previous disk
parameters as a reference point and vary $q\in[-0.2,-1.2]$,
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately.

Here we use a wavenumber $\khat=10$.

% , in which case we find $\nu$
% changes sign as $\beta$ is increased from zero. We find that with
% $\khat=30$, as used in the previous section, when $\beta$ is not small $\nu$ did not reach zero but instead decreases
% rapidly after the expected value of $\beta_\mathrm{crit}$.  These
% modes were found to have large amplitudes near the boundaries.

%we suspect boundary conditions may play a role for

% For large $\khat$ \emph{and} finite
% thermal relaxation (e.g. $\khat=30$, $\beta\gtrsim 0.2$) we found
% $\nu$ did not reach zero, but still decreases rapidly after the
% expected $\beta_\mathrm{crit}$. These modes have large perturbations
% near the boundaries, which may play a role not accounted for in our
% analytical discussion.

The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.

Our numerical results generally agree with 
Eq. \ref{iso_vsi_cond}. The agreement improves with decreasing  $|q|$,
$\epsilon$ and increasing $\gamma$, i.e. weaker instability. 
There is a noticeable difference as $\gamma\to1$ because the
derivation of $\beta_\mathrm{crit}$ assumed $\gamma>1$.
